---
title: "Phenotypic Dataset"
author: "LucianoRogerio"
date: "2021-07-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Wrong section? get back to [home](index.html).

# Phenotypic data - Cassavabase

Initialy we are going to download phenotypic data from Cassavabase.
In this case we are going to download the data from traits related to plant Architecture, as the following table:

<br />

#### Table 1. List of traits downloaded from Cassavabase.

| Trait | Abbreviation | Cassavabase code | Description |
| --- | --- | --- | --- |
| plant architecture visual rating 1-5 | PlantArchitecture | [CO_334:0000099](https://cassavabase.org/cvterm/70722/view) | Plant architecture on a 1-5 scale with 1 = excellent, 2 good, 3 = fair, 4 = bad, and 5 = very bad |
| plant architecture visual rating 1-5 at month 8 | PlantArchitecture8m | COMP:0000119 | Plant architecture on a 1-5 scale with 1 = excellent, 2 good, 3 = fair, 4 = bad, and 5 = very bad |
| flowering ability visual assessment 0-3 | FlowerVisRate | [CO_334:0000233](https://cassavabase.org/cvterm/76963/view) | Presence of flowers (0=none; 1=little; 2=intermediate; 3=many). |
| flower visual rating 0 & 1 | FlowerPresence | [CO_334:0000111](https://cassavabase.org/cvterm/70713/view) | Visual rating of flowers (50%) in plant with 0 = absent and 1 = present. |
| initial plant vigor assessment 1-5 | VigorCIAT | [CO_334:0000220](https://cassavabase.org/cvterm/76955/view) | Visual assessment of plant vigor during establishment (1=very little vigor, and 5 = very vigorous). as being evaluated by CIAT. |
| initial vigor assessment 1-7 | Vigor | [CO_334:0000009](https://cassavabase.org/cvterm/70765/view) | Visual assessment of plant vigor during establishment scored one month after planting. 3 = Not vigorous, 5 = Medium vigor, 7 = highly vigorous. |
| number of forks counting | NFC | [CO_334:0000146](https://cassavabase.org/cvterm/76816/view) | Number of branches (2 forks/branches (dichotomous), 3 forks/branches (trichotomous), or 4 forks/branches (tetrachotomous)) at every branching level. |
| number of forks on branching level 1 counting | NFC1 | [CO_334:0000522](https://cassavabase.org/cvterm/77515/view) | Number of forks (2 forks/branches (dichotomous), 3 forks/branches (trichotomous), or 4 forks/branches (tetrachotomous)) on the first branching level. |
| number of forks on branching level 2 counting | NFC2 | [CO_334:0000523](https://cassavabase.org/cvterm/77519/view) | Number of forks (2 forks/branches (dichotomous), 3 forks/branches (trichotomous), or 4 forks/branches (tetrachotomous)) on the second branching level. |
| number of forks on branching level 3 counting | NFC3 | [CO_334:0000524](https://cassavabase.org/cvterm/77511/view) | Number of forks (2 forks/branches (dichotomous), 3 forks/branches (trichotomous), or 4 forks/branches (tetrachotomous)) on the third branching level. |
| number of nodes at branching level 1 counting | NNC1 | [CO_334:0000352](https://cassavabase.org/cvterm/77487/view) | Number of nodes at the first branching level. |
| number of nodes at branching level 2 counting | NNC2 | [CO_334:0000363](https://cassavabase.org/cvterm/77488/view) | Number of nodes at the second branching level. |
| number of nodes at branching level 3 counting | NNC3 | [CO_334:0000368](https://cassavabase.org/cvterm/77486/view) | Number of nodes at the third branching level. |
| first apical branch height measurement in cm | FirstBranchHeight | [CO_334:0000106](https://cassavabase.org/cvterm/70751/view) | Height of first apical branch (ground level to point of first Apical branch, 9 months after planting) in cm |
| plant height measurement in cm | PlantHeight | [CO_334:0000018](https://cassavabase.org/cvterm/70674/view) | Vertical height of plants from the ground to top of the canopy measured in centimeter (cm). |
| plant height measurement in cm at month 12 | PlantHeight12m | COMP:0000181 | Vertical height of plants from the ground to top of the canopy measured in centimeter (cm). |
| plant height with leaf in cm | PlantHeightLeaf | [CO_334:0000123](https://cassavabase.org/cvterm/76806/view) | Portion of the stem with leaves measured as the distance in centimeter from the point of attachment of the oldest leaf to the youngest leaf (apical leaf portion). |
| plant height without leaf | PlantHeightNLeaf | [CO_334:0000125](https://cassavabase.org/cvterm/76838/view) | Portion of stem with no leaf measured in centimeter (cm) by deducting plant height with leaf from plant height. |
| plant height without leaf at month 12 | PlantHeightNLeaf12m | COMP:0000182 | Portion of stem with no leaf measured in centimeter (cm) by deducting plant height with leaf from plant height. |
| stalk length evaluation | StalkLength | [CO_334:0000227](https://cassavabase.org/cvterm/76941/view) | Visual assessment of the average length of the stalks (1=short; 2=intermediate; 3=long) |
| stem diameter measurement in cm | StemDiam | [CO_334:0000257](https://cassavabase.org/cvterm/77121/view) | Measurement of stem diameter taken on the middle of the plant in centimeter (cm) using the vernier caliper. |
| stem diameter measurement in cm at month 5 | StemDiam5m | COMP:0000129 | Measurement of stem diameter taken on the middle of the plant in centimeter (cm) using the vernier caliper. |
| stem diameter measurement in cm at month 6 | StemDiam6m | COMP:0000130 | Measurement of stem diameter taken on the middle of the plant in centimeter (cm) using the vernier caliper. |
|  |  |  |  |

<br />

## 1. Download of the phenotypic dataset from Cassavabase [wizard tool](https://cassavabase.org/user/login?goto_url=%2Fbreeders%2Fsearch)

Use the following filters:

   1. Select the trait list `TraitsLucianoGS`
   2. Select the following `Trial Types`:
      - *Clonal Evaluation*;
      - *Preliminary Yield Trial*;
      - *Advanced Yield Trial*;
      - *Uniform Yield Trial*;
      - *Regional Trials*;
      - *phenotyping_trial*.
   3. Select `Years` from 2010 to 2021.
   4. Select all the `trials` available at cassavabase after the filter.
   
## 2. Phenotypic data Editing

```{r, eval = FALSE}
library(tidyverse); library(data.table); library(here)

RawPhenoData <- read.csv(file = here::here("data", "phenodata.csv"), header = T, na.strings = "")

RawPhenoData %>% mutate("FlowerPresence" = flower.visual.rating.0.1.CO_334.0000111,
                        "FlowerVisRate" = flowering.ability.visual.assessment.0.3.CO_334.0000233,
                        "VigorCIAT" = initial.plant.vigor.assessment.1.5.CO_334.0000220,
                        "Vigor" = initial.vigor.assessment.1.7.CO_334.0000009,
                        "PlantArchitecture" = plant.architecture.visual.rating.1.5.CO_334.0000099,
                        "PlantHeight" = plant.height.measurement.in.cm.CO_334.0000018,
                        "PlantHeight12m" = plant.height.measurement.in.cm.month.12.COMP.0000181,
                        "PlantHeightLeaf" = plant.height.with.leaf.in.cm.CO_334.0000123,
                        "PlantHeightNLeaf" = plant.height.without.leaf.CO_334.0000125,
                        "PlantHeightNLeaf12m" = plant.height.without.leaf.month.12.COMP.0000182,
                        "StalkLength" = stalk.length.evaluation.CO_334.0000227,
                        "StemDiam" = stem.diameter.measurement.in.cm.CO_334.0000257,
                        "FirstBranchHeight" = first.apical.branch.height.measurement.in.cm.CO_334.0000106,
                     .keep = "unused") %>% melt(data = .,
                                                id.vars = c("studyYear", "programDbId", "programName", "programDescription",
                                                            "studyDbId", "studyName", "studyDesign", "plotWidth",
                                                            "plotLength", "fieldSize", "fieldTrialIsPlannedToBeGenotyped",
                                                            "fieldTrialIsPlannedToCross", "plantingDate", "harvestDate",
                                                            "locationDbId", "locationName", "germplasmDbId", "germplasmName",
                                                            "germplasmSynonyms", "observationLevel", "observationUnitDbId",
                                                            "observationUnitName", "replicate", "blockNumber", "plotNumber",
                                                            "rowNumber", "colNumber", "entryType", "plantNumber"),
                                                variable.name = "Trait",
                                                value.name = "Value") -> PhenoData
PhenoData2 <- PhenoData[!is.na(PhenoData$Value),]

saveRDS(object = PhenoData2, file = here::here("data", "phenotypePAGP.RDS"))
rm(list = ls())

```

## 3. Phenotypic data Information

```{r, message=FALSE, warning=FALSE}
suppressMessages(library(tidyverse)); library(reactable); library(here)
PhenoData <- readRDS(here("data","phenotypePAGP.RDS"))


PhenoData %>% dplyr::select(programName, studyYear, Trait, studyName) %>% unique() -> PhenoData2

Table2 <- table(PhenoData2$programName, PhenoData2$studyYear, PhenoData2$Trait) %>% as.data.frame()
colnames(Table2) <- c("programName", "Year", "Trait", "N˚Trials")
Table2 <- Table2[Table2$`N˚Trials`!=0,]
```

#### Table 2. Number of trial per Institute in Cassavabase with Cassava Plant Shape Traits.
```{r, echo = FALSE}
Table2  %>% reactable(groupBy = c("Trait", "programName"), columns = list(
  "N˚Trials" = colDef(aggregate = "sum")), defaultPageSize = 15)
```
#### Table 3. Plot number of the trials available in Cassavabase with Cassava Plant Shape traits.
```{r, echo = FALSE}
PhenoData %>% dplyr::group_by(Trait, programName, studyYear) %>%
  dplyr::count(studyName) %>% 
   reactable(groupBy = c("Trait", "programName", "studyYear"),
             defaultPageSize = 50,
             columns = list(
               n = colDef(aggregate = "sum", format = colFormat(separators = TRUE))
             ))
```


## 4. Heritability and Reliability analysis per trial

```{r, message = FALSE, warning = FALSE}
library(plyr)
source(here::here("code", "MixedModelsFunctions.R"))

PhenoData$check <- ifelse(PhenoData$entryType!="test", PhenoData$germplasmName, "999") 
PhenoData$new <- ifelse(PhenoData$entryType == "test", 1, 0) 

PhenoData$blockNumber <- as.character(PhenoData$blockNumber)
#PhenoData2 <- PhenoData[!PhenoData$studyName %in% BadTrialsList,]


#fmfit <- PhenoData2 %>% filter(Trait == "PlantArchitecture") %>% dlply(.variables = c("Trait", "studyName"),
#                              .fun = analyzeTrial.sommer)


```
The heritability and Reliability were estimated for each trial.
We will proceed with the trails with reliability and heritability bigger than 0.2 and 0.1, respectively.

Select the trials with unexpected number of blocks, controls for their experimental design
```{r}
PhenoData %>% dplyr::group_by(programName, studyYear, studyDesign, studyName) %>%
  dplyr::summarise(n_blocks = length(unique(blockNumber)),
                   n_reps = length(unique(replicate)),
                   controls = length(unique(check))) -> TrialsInfo

TrialsInfo %>% dplyr::filter((studyDesign %in% c("Augmented", "Alpha", "RCBD") & n_blocks == 1) |
                               (studyDesign == "Augmented" & controls == 1)) -> BadTrialsInfo
BadTrialsList <- BadTrialsInfo$studyName
```


#### Table 4. Number of blocks, replicates and control of the trials available in Cassavabase with Cassava Plant Shape traits.
```{r, echo = FALSE}
TrialsInfo %>% reactable(groupBy = c("programName", "studyYear", "studyDesign"),
                         defaultPageSize = 50)
```

#### Table 5. Trials available in Cassavabase with Cassava Plant Shape traits with strange metadata.
```{r, echo = FALSE}
BadTrialsInfo %>% reactable(groupBy = c("programName", "studyYear", "studyDesign"),
                         defaultPageSize = 50)
```

## Next Steps

- [x] Download Phenotypic dataset;
- [ ] select trials by reliability and heritability;
- [ ] estimate the BLUPS, and genetic correlations between the traits;
- [ ] Create a list with the clone names;
- [ ] Download the Genotypic dataset from the clones phenotyped;
- [ ] Perform the genomic prediction single-trait with G-BLUP Add, Add + Dom genetic models.
      - 50 replicates
      - 100 clones per fold.


[home](index.html)
